---
title: "Bulk TCR repertoire analysis"
output: html_notebook
---




Load libraries

```{r}
library(tcR, quietly = T) # library to process TCR repertoire samples
library(ggseqlogo) # plot logos of sequence motifs
library(pheatmap) # plot (pretty) heatmaps
library(plotly) # interactive plotting
```



# Data processing
Example commands for using MixCR to process repertoire sequencing data
Parallelizing and multithreading these steps is highly advised. Single threaded analysis can run for ~2 hours
More info in: https://github.com/milaboratory/mixcr

```{bash, eval = FALSE}
# align with TCR reference
## alignment can be done only against reference of interest (if testing TCR beta repertoire only align against those)
mixcr align -r sample_TCRA.log -t 10 -f -c TRA -s hs sample_TCRA_1.fastq sample_TCRA_1.fastq sample_TCRA.vdjca
mixcr align -r sample_TCRB.log -t 10 -f -c TRB -s hs sample_TCRB_1.fastq sample_TCRB_1.fastq sample_TCRB.vdjca

# TCR sequence assembly
mixcr assemble -r sample_TCRA.log -t 10 -f sample_TCRA.vdjca sample_TCRA.clns
mixcr assemble -r sample_TCRB.log -t 10 -f sample_TCRA.vdjca sample_TCRB.clns

# export output to a readable table format
mixcr exportClones sample_TCRA.clns sample_TCRA.txt
mixcr exportClones sample_TCRB.clns sample_TCRB.txt
```



# Data preparation
Load data (approx. 5min with this example)

```{r, eval = FALSE}
system("gzip -d test_data/*gz") # unzip test files
cln_txt = list.files(path = "./test_data/", pattern = ".txt") # list all test files
cln_glio_raw = lapply(paste0("./test_data/", cln_txt), 
                       function(x){parse.mixcr(x)}) # parse test files
names(cln_glio_raw) = sapply(strsplit(cln_txt, ".", fixed = T), function(x) x[1])

# get sample metadata from file names
metadata = data.frame(t(sapply(strsplit(names(cln_glio_raw), "_"), rbind)), stringsAsFactors = F)
rownames(metadata) = names(cln_glio_raw)
colnames(metadata) = c("condition", "repertoire", "tissue")

#save(metadata, cln_glio_raw, file = "all_test_data.RData")
```



Load .RData with all needed objects (faster than previous code)

```{r}
load("all_test_data.RData")
```




Out of frame TCR sequences are non functional but can still appear. In https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3872299/: "...out-of-frame (i.e., non-functional and thus not subjected to selection) and in-frame TCR beta repertoires, we also show the extent of the impact of thymic selection and the common trends in how this process shapes individual repertoires.""
Filtering clonotypes in frame/not in frame:

```{r}
cln_glio_list <- get.inframes(cln_glio_raw)
```



# Basic stats about the dataset
Compute basic stats

```{r}
basic_stats = cloneset.stats(cln_glio_list)
basic_stats
```



Stats about clonal proportion

```{r}
# how many clones to make up 25% of the reads?
clonal_prop = clonal.proportion(cln_glio_list, 25)

# proportion taken by the top-10 clonotypes
top_clon_prop = top.proportion(cln_glio_list, 10) 
vis.top.proportions(cln_glio_list)

# clonotype expansion - what proportion of clonotypes fit a certain interval of abundance (as % of total clonotypes)
clonal_space = clonal.space.homeostasis(cln_glio_list)
pheatmap(t(clonal_space), cluster_rows = F)
```



# CDR3 features
CDR3 (nucleotide) length distribution can give us clues about the clonotype diversity. A more skewed distribution implies greater representation of one or a few clonotypes.

```{r}
vis.count.len(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
              .col = "Read.count", .ncol = 2)
vis.count.len(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
              .col = "Read.count", .ncol = 2)
```



Sequence logo for the CDR3 sequence - any evident motifs?
The most correct way to run this analysis would be to have the CDR3 sequences aligned with each other since they can have different lengths. The following code serves mostly as an example on how to plot this in R.

```{r}
plot_list = list()
for(n in names(cln_glio_list)){
  km <- get.kmers(cln_glio_list[[n]]$CDR3.amino.acid.sequence, .head = 100, .k = 16, .verbose = F)
  d <- kmer.profile(km)
  plot_list[[n]] = ggseqlogo(as.matrix(round(d[,-1]*100, 0)), method = 'prob') + ggtitle(n)
  print(plot_list[[n]])
}
```



# Gene segment analysis
## Gene segment usage
Plot gene usage (combinations) in heatmaps
Normalization is required to account for library size, especially when comparing between samples

```{r}
# V segment usage
BV_usage = geneUsage(cln_glio_list, HUMAN_TRBV, .norm = T)
rownames(BV_usage) = BV_usage[,1]
BV_usage = BV_usage[,-1]
BV_usage = BV_usage[,!apply(BV_usage, 2, function(x) all(is.na(x)))]
AV_usage = geneUsage(cln_glio_list, HUMAN_TRAV, .norm = T)
rownames(AV_usage) = AV_usage[,1]
AV_usage = AV_usage[,-1]
AV_usage = AV_usage[,!apply(AV_usage, 2, function(x) all(is.na(x)))]

# J segment usage
BJ_usage = geneUsage(cln_glio_list, HUMAN_TRBJ, .norm = T)
rownames(BJ_usage) = BJ_usage[,1]
BJ_usage = BJ_usage[,-1]
BJ_usage = BJ_usage[,!apply(BJ_usage, 2, function(x) all(is.na(x)))]
AJ_usage = geneUsage(cln_glio_list, HUMAN_TRAJ, .norm = T)
rownames(AJ_usage) = AJ_usage[,1]
AJ_usage = AJ_usage[,-1]
AJ_usage = AJ_usage[,!apply(AJ_usage, 2, function(x) all(is.na(x)))]

both_usage = list()
for(n in names(cln_glio_list)){
  if(metadata[n,"repertoire"]=="TCRB"){
    both_usage[[n]] = geneUsage(cln_glio_list[[n]], list(HUMAN_TRBV, HUMAN_TRBJ), .norm = T)
  } else{
    both_usage[[n]] = geneUsage(cln_glio_list[[n]], list(HUMAN_TRAV, HUMAN_TRAJ), .norm = T)
  }
}

# how are the different samples distinguished by segment usage?
pheatmap::pheatmap(t(BV_usage))
pheatmap::pheatmap(t(AV_usage))
pheatmap::pheatmap(t(BJ_usage))
pheatmap::pheatmap(t(AJ_usage))

# what are the segment combination biases per sample?
pheatmap::pheatmap(t(both_usage$glioma_TCRB_blood), main = "glioma_TCRB_blood")
pheatmap::pheatmap(t(both_usage$glioma_TCRB_brain), main = "glioma_TCRB_brain")
pheatmap::pheatmap(t(both_usage$healthy_TCRB_blood), main = "healthy_TCRB_blood")
pheatmap::pheatmap(t(both_usage$healthy_TCRB_brain), main = "healthy_TCRB_brain")
```



## Gene segment comparison by information distance
Information on diversity measures: https://en.wikipedia.org/wiki/Diversity_index
Entropy - measure of how unpredictable the data is. The larger the entropy, the more diverse (in this case the repertoire).
Compute repertoire entropy

```{r, warning=FALSE}
# V segment usage entropy
BV_entropy = entropy.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], HUMAN_TRBV)
AV_entropy = entropy.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], HUMAN_TRAV)

# J segment usage entropy
BJ_entropy = entropy.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], HUMAN_TRBJ)
AJ_entropy = entropy.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], HUMAN_TRAJ)


# plot
TCRB_df = data.frame(V_seg = BV_entropy,
                     J_seg = BJ_entropy)
TCRB_df = merge(TCRB_df, metadata, by = 0)

ggplot(TCRB_df, aes(x = V_seg, y = J_seg, colour = condition, shape = tissue))+
  geom_point()+
  scale_shape_manual(values = c(2, 19))+
  ggtitle("TCR beta gene segment entropy")+
  theme_classic()+
  theme(aspect.ratio = 1)

TCRA_df = data.frame(V_seg = AV_entropy,
                     J_seg = AJ_entropy)
TCRA_df = merge(TCRA_df, metadata, by = 0)
ggplot(TCRA_df, aes(x = V_seg, y = J_seg, colour = condition, shape = tissue))+
  geom_point()+
  scale_shape_manual(values = c(2, 19))+
  ggtitle("TCR alpha gene segment entropy")+
  theme_classic()+
  theme(aspect.ratio = 1)
```



Other measures of repertoire diversity

```{r}
# True diversity - the most fundamental measure of diversity, a few others are derived from this one
repDiversity(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 'div', 'read.count', 
             .norm = T, .do.norm = T)
repDiversity(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 'div', 'read.count', 
             .norm = T, .do.norm = T)

# Gini coefficient - measure of inequality in the data (the larger the more unequal) (very used in economy)
repDiversity(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 'gini', 'read.prop', 
             .norm = T, .do.norm = T)
repDiversity(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 'gini', 'read.prop', 
             .norm = T, .do.norm = T)
```



Compute divergence between samples
Jensen-Shannon divergence - similarity between two probability distributions. Related to the Kullback–Leibler divergence, a measure of "relative entropy".

```{r}
js_list = list(
  BV_js = js.div.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                     HUMAN_TRBV, .verbose = F),
  AV_js = js.div.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                     HUMAN_TRAV, .verbose = F),
  
  BJ_js = js.div.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                     HUMAN_TRBJ, .verbose = F),
  AJ_js = js.div.seg(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                     HUMAN_TRAJ, .verbose = F)
)

for(n in names(js_list)){pheatmap::pheatmap(js_list[[n]], main = n)}
```



Dimensionality reduction by PCA with gene usage - how do different samples cluster based on their expression of VJ segments?
This case will not be very informative as we are only looking at 4 samples.

```{r}
# use AJ and AV combination
pca_2d = pca.segments.2D(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])],
                              .genes =  list(HUMAN_TRAV, HUMAN_TRAJ), .do.plot = F)

plot_df = merge(pca_2d$x, metadata, by = 0)

ggplot(plot_df, aes(x = PC1, y = PC2, colour = condition, shape = tissue))+
  geom_point(size = 2)+
  scale_shape_manual(values = c(19, 2))+
  ggtitle("TCR alpha PCA")+
  theme_classic()+
  theme(aspect.ratio = 1,
        legend.position = "right")

# use BJ and BV combination
pca_2d = pca.segments.2D(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])],
                              .genes =  list(HUMAN_TRBV, HUMAN_TRBJ), .do.plot = F)

plot_df = merge(pca_2d$x, metadata, by = 0)

ggplot(plot_df, aes(x = PC1, y = PC2, colour = condition, shape = tissue))+
  geom_point(size = 2)+
  scale_shape_manual(values = c(19, 2))+
  ggtitle("TCR beta PCA")+
  theme_classic()+
  theme(aspect.ratio = 1,
        legend.position = "right")
```



# Comparing repertoires
Compute a pairwise normalised number of shared clonotypes and plot a heatmap of them
Normalisation is important to avoid bisases by different library size and clonotype capture

```{r}
# TCRA
ovelap_A_nuc_exact <- repOverlap(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                                    "exact", .norm = T, .seq = "nuc")
ovelap_A_aa_exact <- repOverlap(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                                   "exact", .norm = T, .seq = "aa")
vis.heatmap(ovelap_A_nuc_exact, .text = F, .title = "TCRA clonotype overlap\nNucleotide Sequence")
vis.heatmap(ovelap_A_aa_exact, .text = F, .title = "TCRA clonotype overlap\nAminoacid Sequence")

# TCRB
ovelap_B_nuc_exact <- repOverlap(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                                    "exact", .norm = T, .seq = "nuc")
ovelap_B_aa_exact <- repOverlap(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                                   "exact", .norm = T, .seq = "aa")
vis.heatmap(ovelap_B_nuc_exact, .text = F, .title = "TCRB clonotype overlap\nNucleotide Sequence")
vis.heatmap(ovelap_B_aa_exact, .text = F, .title = "TCRB clonotype overlap\nAminoacid Sequence")
```



Top cross - number of shared clonotypes among those most highly represented
Important because these tend to be the most relevant forms of the TCR

```{r}
tcra_topcross <- top.cross(.data = cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                     .n = seq(500, 10000, 500), .verbose = F, .norm = T)
top.cross.plot(tcra_topcross)

tcrb_topcross <- top.cross(.data = cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                     .n = seq(500, 10000, 500), .verbose = F, .norm = T)
top.cross.plot(tcrb_topcross)
```



Shared repertoire - number of shared clonotypes for each repertoire for each degree of sharing (i.e., number of samples in which indicated amount of clones have been found).

```{r}
shared_A_rep = shared.repertoire(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                                 .min.ppl = 2, .clear = T)
shared.representation(shared_A_rep)

shared_B_rep = shared.repertoire(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                                 .min.ppl = 2, .clear = T)
shared.representation(shared_B_rep)
```



Shared repertoire mutation networks - how do the different clonotypes relate to each other by sequence similarity?

```{r}
# TCRBA
shared_A_rep_top = shared.repertoire(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRA",])], 
                                     .min.ppl = 2, .clear = T, .head = 2000)
G_A <- mutation.network(shared_A_rep_top, .max.errors = 3)

## plotting attributes
V(G_A)$color = metadata[metadata$repertoire=="TCRA","condition"]
V(G_A)$shape = metadata[metadata$repertoire=="TCRA","tissue"]

## network layout
l_A = igraph::layout.fruchterman.reingold(G_A)

# TCRB
shared_B_rep_top = shared.repertoire(cln_glio_list[rownames(metadata[metadata$repertoire=="TCRB",])], 
                                     .min.ppl = 2, .clear = T, .head = 2000)
G_B <- mutation.network(shared_B_rep_top, .max.errors = 3)

## plotting attributes
V(G_B)$color = metadata[metadata$repertoire=="TCRB","condition"]
V(G_B)$shape = metadata[metadata$repertoire=="TCRB","tissue"]

## network layout
l_B = igraph::layout.fruchterman.reingold(G_B)
```



Making the actual network plot
More info: https://plot.ly/r/network-graphs/

```{r}
# ploting function
makePlotlyNetwork = function(igraph_obj, layout_net, title_net){
  vs <- V(igraph_obj)
  es <- as.data.frame(get.edgelist(igraph_obj))
  
  Nv <- length(vs)
  Ne <- length(es[1]$V1)
  
  Xn <- layout_net[,1]
  Yn <- layout_net[,2]
  
  network <- plotly::plot_ly(x = ~Xn, y = ~Yn, mode = "markers", text = vs$label, 
                             hoverinfo = "text", color = vs$color, symbol = vs$shape, 
                             symbols = c("19", "2"), colors = c("#FF0006", "#0080FF"))
  edge_shapes <- list()
  for(i in 1:Ne) {
    v0 <- es[i,]$V1
    v1 <- es[i,]$V2
  
    edge_shape = list(
      type = "line",
      line = list(color = "#030303", width = 0.3),
      x0 = Xn[v0],
      y0 = Yn[v0],
      x1 = Xn[v1],
      y1 = Yn[v1]
    )
  
    edge_shapes[[i]] <- edge_shape
  }
  
  axis <- list(title = "", showgrid = FALSE, showticklabels = FALSE, zeroline = FALSE)
  
  p <- plotly::layout(
    network,
    title = title_net,
    shapes = edge_shapes,
    xaxis = axis,
    yaxis = axis
  )
}

# make the plots
p = makePlotlyNetwork(G_A, l_A, "TCRA mutation graph")
p
p = makePlotlyNetwork(G_B, l_B, "TCRB mutation graph")
p
```







